%
% The Mite virtual machine
%
% Reuben Thomas   20/5/00-24/6/01
%

% Work out what the possible error conditions are in verified code & list them.


\documentclass[english]{scrartcl}
% Put usepackages one per line for HEVEA
\usepackage[LY1]{fontenc}
\usepackage{babel}
\usepackage{lypslatex}
\usepackage{array}
\usepackage{amstext}
\usepackage{ctable}
\usepackage{booktabs}
\usepackage{url}

% Settings for this document

\makeatletter
\urlstyle{rm}
\def\UrlSpecials{\do\<{\langle}\do\>{\mathbin{\rangle}}\do\_{\_%
 \penalty\@m}\do\|{\mid}\do\{{\lbrace}\do\}{\mathbin{\rbrace}}\do
 \\{\char47\char47}%
 \do\~{\raise.6ex\hbox{\m@th$\scriptstyle\sim$}}%
 \do \ {\ }}%
\makeatother

% Macros for this document
\newcommand{\cont}{\noindent}
\newcommand{\synfont}{\sffamily}
\newcommand{\syn}[1]{{\synfont #1}}
\newcommand{\norm}[1]{\textrm{#1}}

% Instruction tables
\newenvironment{insttab}[1]
    {\begin{ctable}{#1}{>{\synfont}p{13ex}p{44ex}}\toprule}
    {\bottomrule\end{ctable}}


\begin{document}

\title{The Mite virtual machine}
\author{Reuben Thomas\\\url{rrt@sc3d.org}\\\url{http://sc3d.org/rrt/}}
\date{29th April 2001}
\maketitle



\section{Introduction}

Mite is a general-purpose virtual machine. It is designed to be capable of
efficient implementation, by interpretation or compilation, from a
binary-portable object format. It has a flat linear memory, and
single-threaded, non-interruptable execution. The only external access
provided is the ability to call machine code.



\section{Parameters}

Mite is parametrized on the following quantities:

\begin{description}
\item[$w$]number of bytes in a word

\item[$g$]the number of general-purpose registers ($g=2^n, 2<n$, for some $n$)

\item[$s$]direction of stack growth ($-1$ for an ascending stack, $1$ for a
descending stack)
\end{description}



\section{Registers and memory}

Registers are word-sized. Mite has $g$ general registers \syn{1} to $g$, a
program counter \syn{P}, and a stack pointer \syn{S}. The stack is a LIFO
stack; \syn{S} points to the top-most item.

The memory $M$ is an array of bytes. The stack resides in memory. $M(a)$
denotes the word in memory starting at byte $a$; $a$ must be a multiple of
$w$. Within a word, bytes may be stored in big or little-endian order. Two's
complement number representation is used.



\section{Execution}

Mite repeatedly loads the instruction at \syn{P}, makes \syn{P} point to
the next instruction, and executes the loaded instruction. Program addresses
need not be addresses in $M$.



\section{Instructions}

In the syntax, each operand is denoted by a letter subscripted by a number.
The letter indicates the sort of operand required, as shown in
table~\ref{optypetab}; the number shows the number of the operand. A name is
a string of letters, digits and underscores, and may not start with a digit.

\begin{ctable}{Operand types\label{optypetab}}{>{$}c<{$}l}
\toprule
r & a general register\\
l & a label type\\
i & an immediate constant\\
n & a name\\
\bottomrule
\end{ctable}

$\{P\}$ has the value $1$ if the predicate $P$ is true, and $0$ otherwise.


\subsection{Labels}

Labels are defined by the \syn{lab} instruction, shown in table~\ref{labtab}.

\begin{insttab}{Label definition\label{labtab}}
lab $l_1$ $n_2$        & define a label\\
\end{insttab}

There are three types of label: branch labels, denoted \syn{b}, subroutine
labels, denoted \syn{s}, and data labels, denoted \syn{d}. Names must be
unique within a translation unit.

The value of a code or subroutine label is the address of the instruction
immediately following it. A data label's value is the address of the first
datum stored after it by a data instruction.

Branch labels have no effect when they are executed; subroutine and data
labels may not be executed.


\subsection{Computation}

\begin{insttab}{Computational instructions\label{comptab}}
mov $r_1$ $r_2$        & $r_1\gets r_2$ \\
movi $r_1$ $i_2$       & $r_1\gets i_2$ \\
ldl $r_1$ $n_2$        & $r_1\gets n_2$ \\
ld $r_1$ $r_2$         & $r_1\gets M(r_2)$ \\
st $r_1$ $r_2$         & $M(r_2)\gets r_1$ \\
gets $r_1$             & $r_1\gets S$ \\
sets $r_1$             & $S\gets r_1$ \\
pop $r_1$              & $r_1\gets M(S)$; $S\gets S-sw$ \\
push $r_1$             & $S\gets S+sw$; $M(S)\gets r_1$ \\
add $r_1$ $r_2$ $r_3$  & $r_1\gets r_2+r_3$ \\
sub $r_1$ $r_2$ $r_3$  & $r_1\gets r_2-r_3$ \\
mul $r_1$ $r_2$ $r_3$  & $r_1\gets r_2\times r_3$ \\
div $r_1$ $r_2$ $r_3$  & $r_1\gets r_2\div r_3$ (unsigned) \\
rem $r_1$ $r_2$ $r_3$  & $r_1\gets r_2\bmod r_3$ (unsigned) \\
and $r_1$ $r_2$ $r_3$  & $r_1\gets r_2$ bitwise and $r_3$ \\
or $r_1$ $r_2$ $r_3$   & $r_1\gets r_2$ bitwise or $r_3$ \\
xor $r_1$ $r_2$ $r_3$  & $r_1\gets r_2$ bitwise xor $r_3$ \\
sl $r_1$ $r_2$ $r_3$   & $r_1\gets r_2$ left shift $r_3$ ($0\leq r_3\leq 8w$) \\
srl $r_1$ $r_2$ $r_3$  & $r_1\gets r_2$ logical right shift $r_3$ ($0\leq r_3\leq 8w$) \\
sra $r_1$ $r_2$ $r_3$  & $r_1\gets r_2$ arithmetic right shift $r_3$ ($0\leq r_3\leq 8w$) \\
teq $r_1$ $r_2$ $r_3$  & $r_1\gets\{r_2=r_3\}$ \\
tlt $r_1$ $r_2$ $r_3$  & $r_1\gets\{r_2<r_3\}$ \\
tltu $r_1$ $r_2$ $r_3$ & $r_1\gets\{r_2<r_3$ (unsigned)$\}$ \\
b $n_1$                & $P\gets n_1$ \\
br $r_1$               & $P\gets r_1$ \\
bf $r_1$ $n_2$         & if $r_1=0$, $P\gets n_2$ \\
bt $r_1$ $n_2$         & if $r_1\neq0$, $P\gets n_2$ \\
call $n_1$             & $S\gets S+sw$; $M(S)\gets P$; $P\gets n_1$ \\
callr $r_1$            & $S\gets S+sw$; $M(S)\gets P$; $P\gets n_1$ \\
ret                    & $P\gets M(S)$; $S\gets S-sw$ \\
calln $r_1$            & call native code at $r_1$ \\
\end{insttab}

The computational instructions are shown in table~\ref{comptab}.

Immediate constants are of the form
[\syn{e}][\syn{s}][\syn{w}]$n$[\syn{>\/>}$r$]. $n$ (and $r$, if present) is
an integer. If \syn{e} (for ``endianness'') is present, $n$ is subtracted
from $w$ on a big-endian Mite, or left unaltered otherwise; then if \syn{s}
is present, it is multiplied by $s$, and if \syn{w} is present, it is
multiplied by $w$. The final value is truncated to $8w$ bits, then if
\syn{>\/>}$r$ is present, it is rotated by $r$ places, to the right if
positive, and to the left if negative.


\subsection{Data}

The data instructions, shown in table~\ref{datatab}, allow literal data to
be included in object code; they may not be executed. Data between two
labels are stored contiguously in $M$.

\begin{insttab}{Data instructions\label{datatab}}
lit $i_1$              & a literal word \\
litl $l_1$ $n_2$       & a literal label \\
space $i_1$            & $i_1$ zero words ($i_1>0$)\\
\end{insttab}

\syn{lit} causes its operand to be truncated to a word and stored in the
next word of memory. \syn{litl} causes the values of the given label to be
stored in the next word of memory.

\syn{space} causes the given number of zero words to be stored in
consecutive locations.



\section{Object format}

In the description below, hexadecimal numbers are indicated by a trailing
``h''.

Object code consists of a series of instructions.


\begin{ctable}{Instruction opcodes\label{opcodetab}}{*{3}{>{\synfont}cc!{\hspace{5mm}}}}
\toprule
\norm{\bf Instruction} & \bf Opcode & \norm{\bf Instruction} & \bf Opcode &
\norm{\bf Instruction} & \bf Opcode \\ \midrule
lab   & 01h & mul   & 0Dh & b     & 19h \\
mov   & 02h & div   & 0Eh & br    & 1Ah \\
movi  & 03h & rem   & 0Fh & bf    & 1Bh \\
ldl   & 04h & and   & 10h & bt    & 1Ch \\
ld    & 05h & or    & 11h & call  & 1Dh \\
st    & 06h & xor   & 12h & callr & 1Eh \\
gets  & 07h & sl    & 13h & ret   & 1Fh \\
sets  & 08h & srl   & 14h & calln & 20h \\
pop   & 09h & sra   & 15h & lit   & 21h \\
push  & 0Ah & teq   & 16h & litl  & 22h \\
add   & 0Bh & tlt   & 17h & space & 23h \\
sub   & 0Ch & tltu  & 18h \\
\bottomrule
\end{ctable}

\subsection{Instructions} \label{objinst}

Instructions are encoded as the opcode followed by the operands, in
numerical order. Instructions that do not end on a $4$-byte boundary are
padded to the next such boundary with zero bytes.

Opcodes, registers and label types are encoded as one byte. The instruction
opcodes are shown in table~\ref{opcodetab}. Opcodes 25h--7Fh are reserved
for future expansion. Registers are encoded by their number. Label types are
encoded as $1$ for \syn{b}, $2$ for \syn{s}, and $3$ for \syn{d}.

Labels are encoded as a long number (see section~\ref{longnums}). Labels of
each type are numbered consecutively from $1$ from the start of the
translation unit, according to the order in which they are declared. For a
\syn{lab} instruction, only the label type is encoded (the number is
redundant).

Immediate constants are encoded as a byte encoding the modifiers followed by
a byte encoding the rotation (if any), followed by the basic value encoded
as a long number. The modifiers are encoded as binary flags, as shown in
table~\ref{modenctab}; the unused bits are zeroed.

\begin{ctable}{Constant modifiers encoding\label{modenctab}}{>{\synfont}cc}
\toprule
\bf Modifier & \bf Bit position \\ \midrule
r & 0 \\
w & 1 \\
s & 2 \\
e & 3 \\
\bottomrule
\end{ctable}


\subsection{Long numbers} \label{longnums}

Long numbers are encoded as follows:

\begin{enumerate}
\item Left-truncate the number to the minimum number of bits in which it can
be represented.

\item Remove an $8n-1$-bit word from the most significant end of the number,
where $n$ is the number of bytes left in the current instruction word. If
there are fewer than $8n-1$ bits in the number then sign-extend it to that
length first.

\item Add a bit to the most significant end of the word, which should be
zero if more bits remain in the number, and one otherwise.

\item Store the word, with the bytes in big-endian order.

\item While there are still bits in the number, repeat the following steps:

\begin{enumerate}

\item Remove a $31$-bit word from the most significant end of the number,
padding with zeroes if there are fewer than $31$ bits left.

\item Add a bit to the most significant end of the word, which should be
zero if more bits remain in the number, and one otherwise.

\item Store the word, with the bytes in big-endian order.
\end{enumerate}

\end{enumerate}



\section{Acknowledgements}

Martin Richards introduced me to Cintcode~\cite{cintcode}, which kindled my
interest in virtual machines, and led to Beetle~\cite{beetledis} and an
earlier version of Mite~\cite{mite}, of which the current Mite is a sort of
synthesis. GNU~\emph{lightning}~\cite{lightning} helped inspire me to
greater simplicity, while still aiming for good performance. Julian Seward
helped me think about Mite's usefulness as a compiler target, and
interpretive performance. Alistair Turnbull has been a fount of criticism
for all my work on virtual machines.


\bibliographystyle{plain}
\bibliography{rrt,vm}

\end{document}
